---
title: "BirdNET species thresholds: methods report (first draft)"
author: "Alban Sagouis"
date: "2026-10-08"
output:
  github_document:
    html_preview: false
    md_extensions: -implicit_figures
    pandoc_args: ["--markdown-headings=atx"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)

source(file = here::here("R", "birds", "config.R"))

# Every number in this report is read from data/processed/ and outputs/ (written
# by R/birds/01 to 04) or from the raw predictions. Nothing is typed by hand.
read_processed <- function(file) {
  readr::read_csv(
    file = here::here("data", "processed", file),
    show_col_types = FALSE
  )
}

evidence <- read_processed(file = "birds_threshold_evidence.csv")
aic <- read_processed(file = "birds_aic.csv")
calibration <- read_processed(file = "birds_calibration.csv")
coverage <- read_processed(file = "birds_covariate_coverage.csv")
daypart <- read_processed(file = "birds_daypart.csv")
device_boot <- read_processed(file = "birds_device_bootstrap.csv")
leave_device_out <- read_processed(file = "birds_leave_device_out.csv")
link_diag <- read_processed(file = "birds_link_diagnostics.csv")
min_group <- read_processed(file = "birds_min_group_sensitivity.csv")
sensitivity <- read_processed(file = "birds_threshold_sensitivity.csv")
linked <- read_processed(file = "birds_validations_linked.csv")

# Raw predictions (read only, never modified): total and score medians
predictions <- readr::read_csv(
  file = birds_config$paths$predictions,
  col_select = c("common_name", "confidence"),
  show_col_types = FALSE
)
n_predictions <- nrow(predictions)

# Observation labels written by R/birds/04_label_observations.R
labelled <- readr::read_csv(
  file = birds_config$paths$labelled,
  col_select = c("common_name", "observation_status"),
  show_col_types = FALSE
)

# Formatting helpers
f1 <- function(x) sprintf(fmt = "%.1f", x)
f2 <- function(x) sprintf(fmt = "%.2f", x)
f3 <- function(x) sprintf(fmt = "%.3f", x)
pct0 <- function(x) sprintf(fmt = "%.0f%%", 100 * x)
pct1 <- function(x) sprintf(fmt = "%.1f%%", 100 * x)
big <- function(x) format(x, big.mark = ",", trim = TRUE)

# One row per species, and one-row look-ups used in the prose
sp_nightjar <- "Abyssinian Nightjar"
sp_firefinch <- "Red-billed Firefinch"
sp_oriole <- "African Black-headed Oriole"
sp_plover <- "Three-banded Plover"
pick <- function(data, species) filter(data, commonName == species)

nj <- pick(data = evidence, species = sp_nightjar)
ff <- pick(data = evidence, species = sp_firefinch)
pl <- pick(data = evidence, species = sp_plover)
ori <- pick(data = evidence, species = sp_oriole)

# Median score of the validated clips against all predictions
median_score <- function(data, species, column) {
  data |>
    filter(.data[[column]] == species) |>
    pull(confidence) |>
    median()
}
nj_med_val <- median_score(data = linked, species = sp_nightjar, column = "commonName")
nj_med_pred <- median_score(data = predictions, species = sp_nightjar, column = "common_name")
ff_med_val <- median_score(data = linked, species = sp_firefinch, column = "commonName")
ff_med_pred <- median_score(data = predictions, species = sp_firefinch, column = "common_name")

n_validated <- sum(evidence$n_neg + evidence$n_pos)
n_fitted <- sum(evidence$status == "fitted threshold")

# Input checks: the prose assumes one fitted species, the nightjar
checkmate::assert_true(x = n_fitted == 1L)
checkmate::assert_true(x = nj$status == "fitted threshold")
checkmate::assert_true(x = n_validated == nrow(linked))
checkmate::assert_true(x = nrow(labelled) == n_predictions)
checkmate::assert_set_equal(
  x = evidence$commonName,
  y = c(sp_nightjar, sp_oriole, sp_firefinch, sp_plover)
)

# Linking
link_exact <- filter(link_diag, rule == "exact recording name, rounded score")
link_final <- filter(link_diag, stringi::stri_detect_fixed(str = rule, pattern = "(final)"))
n_drift <- sum(!stringi::stri_endswith_fixed(str = linked$recording, pattern = "00"))
checkmate::assert_true(x = link_exact$n_validations == n_validated)

# Clamp sensitivity (nightjar), rounded to three decimals
nj_sens <- pick(data = sensitivity, species = sp_nightjar)
nj_clamp <- nj_sens |>
  filter(parameter == "clamp_max") |>
  pull(glm_threshold) |>
  round(digits = 3)
checkmate::assert_true(x = dplyr::n_distinct(nj_clamp) == 1L)

# Recorders (device) coverage
dev_summary <- function(species, what) {
  coverage |>
    filter(commonName == species, dimension == "summary", level == what) |>
    pull(value)
}
nj_devices <- coverage |>
  filter(commonName == sp_nightjar, dimension == "device") |>
  arrange(desc(n_neg))
nj_top2 <- nj_devices[[1, "level"]] |> c(nj_devices[[2, "level"]])
checkmate::assert_true(x = all(nj_devices$n_pos[1:2] == 0L))

# Leave-one-recorder-out
lodo_single <- function(species) {
  leave_device_out |>
    filter(
      commonName == species,
      !stringi::stri_detect_fixed(str = dropped, pattern = "none"),
      !stringi::stri_detect_fixed(str = dropped, pattern = "+")
    )
}
nj_lodo <- lodo_single(species = sp_nightjar)
ff_lodo <- lodo_single(species = sp_firefinch)
nj_lodo_pair <- leave_device_out |>
  filter(
    commonName == sp_nightjar,
    stringi::stri_detect_fixed(str = dropped, pattern = "+")
  )
checkmate::assert_true(x = nrow(nj_lodo_pair) == 1L)

nj_dboot <- pick(data = device_boot, species = sp_nightjar)
ff_dboot <- pick(data = device_boot, species = sp_firefinch)

# Day-part
dp_nj <- pick(data = daypart, species = sp_nightjar)
dp_ff <- pick(data = daypart, species = sp_firefinch)
dp_nj_night <- filter(dp_nj, n_neg == 0L)
dp_ff_dawn <- filter(dp_ff, stringi::stri_startswith_fixed(str = daypart, pattern = "dawn"))

# Calibration, top nightjar band
cal_top <- calibration |>
  filter(commonName == sp_nightjar, is_top_band)

# Recording period (all validated clips)
rec_dates <- as.Date(
  x = stringi::stri_sub(str = linked$recording, from = 7L, to = 14L),
  format = "%Y%m%d"
)
period <- range(rec_dates)
n_days <- as.integer(diff(period)) + 1L

# Nightjar by week of the recording period, and by the two recorders holding
# most of the wrong clips (potential date effect)
fmt_day <- function(d) sub(pattern = "^0", replacement = "", x = format(d, "%d %B"))
nj_clips <- linked |>
  filter(commonName == sp_nightjar) |>
  mutate(
    device = stringi::stri_extract_first_regex(str = recording, pattern = "^[A-Za-z0-9]+"),
    date = as.Date(stringi::stri_sub(str = recording, from = 7L, to = 14L), format = "%Y%m%d"),
    week = as.integer(floor(as.numeric(date - min(rec_dates)) / 7)) + 1L,
    top2 = is.element(el = device, set = nj_top2)
  )
nj_week <- nj_clips |>
  summarise(
    clips = n(),
    wrong = sum(outcome == 0L),
    wrong_top2 = sum(outcome == 0L & top2),
    clips_other = sum(!top2),
    wrong_other = sum(outcome == 0L & !top2),
    .by = week
  ) |>
  arrange(week)
# Same weekly clip counts as the coverage table
nj_week_cov <- coverage |>
  filter(commonName == sp_nightjar, dimension == "week") |>
  arrange(level)
checkmate::assert_true(x = all(nj_week$clips == nj_week_cov$n_clips))
nj_late <- filter(nj_week, week >= 2L)
nj_top2_dates <- nj_clips |>
  filter(top2) |>
  pull(date) |>
  range()
nj_other_by_week <- nj_week |>
  filter(clips_other > 0L) |>
  mutate(text = paste0(wrong_other, " of ", clips_other, " in week ", week)) |>
  pull(text)
```

## 1. TL;DR

- **What I did.** I followed the Wood & Kahl (2024) recipe to turn
  `r n_validated` expert-validated BirdNET clips into a confidence
  threshold per species, for four species, and added checks for model
  fit, recorder effects and day-part effects (#3).
- **Key result.** Only the Abyssinian Nightjar gets a threshold:
  `r f3(nj$glm_threshold)` (bootstrap interval `r f2(nj$boot_lower)` to
  `r f2(nj$boot_upper)`). The other three (African Black-headed Oriole,
  Three-banded Plover, Red-billed Firefinch) get `no_threshold` because
  the validation sample cannot support a fit.
- **Key decision.** One method and one precision target
  (`r birds_config$target_precision`) for every species. A species that
  fails the data gates is not given a special-case threshold. It is
  given a request for more validation.
- **What is uncertain.** The `r birds_config$target_precision` is a
  claim from the model, not something observed. The nightjar threshold
  rests on `r nj$n_neg` wrong clips, almost all from two recorders and
  all from the evening, so its reach beyond those conditions is not
  established.
- **What I ask of the experts.** Random validation clips across more
  areas, periods and recorders for every species, and especially random
  nightjar clips at night (section 9).

## 2. Data and how it was validated

BirdNET gave `r big(n_predictions)` predictions. The experts validated
`r n_validated` clips across four species. Each clip was marked right
(outcome 1) or wrong (outcome 0). I call these right clips and wrong
clips throughout.

```{r chunk-validated}
evidence |>
  arrange(commonName) |>
  mutate(
    score_min = f3(score_min),
    score_max = f3(score_max)
  ) |>
  select(
    Species = commonName,
    `Wrong clips` = n_neg,
    `Right clips` = n_pos,
    `Lowest score` = score_min,
    `Highest score` = score_max
  ) |>
  knitr::kable(format = "pipe", align = c("l", "r", "r", "l", "l"))
```

### 2.1 The validation sample is enriched, not random

Wood & Kahl step 1 asks for validation clips spread across the whole
confidence range, so that the curve has data at both ends. This is also
why the share of right clips in the sample is not the precision of
BirdNET in the full data.

I do not know how the clips were chosen. They score higher than the
predictions as a whole (nightjar median `r f2(nj_med_val)` against
`r f2(nj_med_pred)`), which fits a sample enriched for high scores. The
firefinch is the exception (`r f2(ff_med_val)` against
`r f2(ff_med_pred)`). If the clips were also picked for being clear
examples, the fitted curve would look better than it should, and I
cannot rule that out.

### 2.2 Linking validations to predictions

The validation file has no prediction ID, so I link on recording,
species and score as a best effort and report how well it works.
Nothing in the thresholds or the labels depends on that link. Matching
on the exact recording name leaves `r link_exact$unmatched` of
`r n_validated` validations unmatched (only `r n_validated - n_drift` of
the `r n_validated` recording names are found verbatim in the
predictions). The cause is recorder clock drift: `r n_drift` validation
recording names have a seconds field other than `00`, while the
predictions always end in `00`. Snapping start times to the nearest
hour, and comparing scores at three decimals, gives
`r link_final$unique` unique matches, `r link_final$ambiguous` ambiguous
and `r link_final$unmatched` unmatched. Of those unmatched,
`r link_final$residual_within_0.001` are rounding differences under
0.001 and `r link_final$residual_recording_absent` recording is absent
from the predictions.

The missing prediction ID is an upstream data issue. The fix is for the
validation export to carry the clip's start position in the recording
(or a prediction ID), the same recording identifier as the predictions,
and scores at full precision. The handoff to Tech (#5) spells this out.

## 3. The model

For each species I fit a logistic regression of the clip outcome (right
or wrong) on the BirdNET confidence, after transforming the confidence
to the logit scale (the log of odds, which stretches the 0 to 1 range so
scores near 1 are not squeezed together). This is the Wood & Kahl
method. The fitted curve gives the chance that a prediction is right at
any score. The threshold is the score at which that chance reaches
`r birds_config$target_precision`, found by solving the fitted curve for
`r birds_config$target_precision` (`r birds_config$target_precision` is
set in the config file as the target precision).

### 3.1 One addition of mine

A confidence of exactly 1.0 has no finite logit, and the Wood & Kahl
code does not handle it. I set those scores to
`r birds_config$clamp_max` before the transform. Moving this value to
`r birds_config$clamp_sensitivity[[1]]` or `r birds_config$clamp_sensitivity[[3]]`
does not change the nightjar threshold at three decimals
(`r f3(nj_clamp[[1]])` in all three cases).

### 3.2 What the confidence column is

The confidence column is the bounded 0 to 1 BirdNET score. The BirdNET
sensitivity setting is not recorded in the data, but it does not change
the threshold in confidence units.

## 4. Who gets a threshold

Two separate questions.

### 4.1 Question 1: do the data allow a threshold?

This decides whether a species gets one. There are two gates, applied
identically to every species:

1.  The fitted threshold lies inside the range of scores that were
    validated.
2.  The smaller of the two outcome groups (wrong clips or right clips)
    has at least `r birds_config$min_smaller_group` clips.

The `r birds_config$min_smaller_group` is deliberately strict. It is set
in the config file and could be relaxed if expert validation time makes
it unreasonable. The table shows what happens at
`r birds_config$min_smaller_group_sensitivity[[1]]`,
`r birds_config$min_smaller_group_sensitivity[[2]]` and
`r birds_config$min_smaller_group_sensitivity[[3]]`.

```{r chunk-min-group}
min_group |>
  select(
    `Minimum in smaller group` = min_smaller_group,
    `Species passing` = n_pass,
    Which = species_pass
  ) |>
  knitr::kable(format = "pipe", align = c("r", "r", "l"))
```

The answer is the same at `r birds_config$min_smaller_group_sensitivity[[1]]`,
`r birds_config$min_smaller_group_sensitivity[[2]]` and
`r birds_config$min_smaller_group_sensitivity[[3]]`: only the nightjar
passes (smaller group of `r nj$smaller_group`). The firefinch has a
smaller group of `r ff$smaller_group`, so it would pass the count gate
at `r birds_config$min_smaller_group_sensitivity[[1]]`, but its fitted
threshold (`r f3(ff$glm_threshold)`) sits above every validated score
(highest `r f3(ff$score_max)`), so it fails the range gate. Plover
(`r pl$n_neg` wrong clip) and oriole (`r ori$n_neg` wrong clips) fail the
count gate at every setting.

Stability under bootstrap is reported, not a gate. The bootstrap
(resampling with replacement) redraws the validated clips, as many as
the original sample, and refits the curve `r big(birds_config$n_boot)`
times. I report the share of bootstrap fits that fail and the interval of
the bootstrapped thresholds. Today the count and range gates already give
the same answer, so a third gate would add a number to defend without a
new result. I keep the numbers because refits will need them (section 8).

### 4.2 Question 2: does it generalise?

A threshold is fitted on the clips the experts validated. Question 2
asks whether it would still hold for clips from other recorders, other
hours of the day, other dates or other places.

This question does not decide whether a species gets a threshold: only
the two gates in 4.1 do. Instead, the answer is reported next to the
threshold as its **scope**, meaning the conditions it was actually
checked under. The checks are in section 7. For the nightjar they show
that the threshold rests mostly on evening clips from a few recorders,
so its reach beyond those conditions is not established.

### 4.3 One method and one precision target for every species

No species gets a stricter or looser cutoff than the method gives. More
certainty is available (more validation, a more cautious cutoff), but it
costs expert time and cannot reasonably be afforded for every species. A
stricter cutoff for some species would lower their counts for reasons
unrelated to biology and distort relative abundances across species. One
limit to state: a common precision target does not make recall equal
across species. Wood & Kahl note this and say recall may matter more for
between-species comparisons. I have no recall figure in the data, so I
do not report one here.

## 5. Results per species

```{r chunk-results}
evidence |>
  mutate(
    estimate = if_else(is.na(glm_threshold), "none", f3(glm_threshold)),
    interval = if_else(
      is.na(boot_lower),
      "none",
      paste(f2(boot_lower), "to", f2(boot_upper))
    ),
    failed = pct0(boot_share_failed),
    outcome = if_else(status == "fitted threshold", status, "no_threshold")
  ) |>
  select(
    Species = commonName,
    `Wrong clips` = n_neg,
    `Right clips` = n_pos,
    `Smaller group` = smaller_group,
    `In range` = in_range,
    `Threshold estimate` = estimate,
    `Bootstrap interval` = interval,
    `Failed bootstrap fits` = failed,
    Outcome = outcome
  ) |>
  knitr::kable(format = "pipe", align = c("l", "r", "r", "r", "l", "l", "l", "l", "l"))
```

The threshold estimate is shown for every species the model could fit,
for transparency. It becomes a threshold only where the outcome says so.
The plover estimate comes from a single wrong clip and the firefinch
estimate lies outside the validated range.

The nightjar threshold is `r f3(nj$glm_threshold)`. In the validation
sample, `r nj$prec_at_glm_n_pos` of the `r nj$prec_at_glm_n` clips at or
above it were right. With so few clips, the lower end of a
`r birds_config$ci_level * 100`% interval on that precision is
`r f3(nj$prec_at_glm_cp_lower)`, so the data do not rule out a real
precision below `r birds_config$target_precision`.

![Validated clips and the fitted curve per species. The red line marks the threshold, fitted species only; the dashed line is a 0.99 chance of being correct. Description: Four panels, one per species, showing validated clips as right (top) or wrong (bottom) against BirdNET confidence. Only the Abyssinian Nightjar panel has a fitted curve and a vertical threshold line near 0.67.](figures/birds_threshold_fits.png)

Labelling uses the point estimate: predictions at or above the fitted
threshold are `observed`. The bootstrap interval is stored and reported
next to it, but there is no second, more cautious cutoff.

## 6. Model checks

### 6.1 Calibration

The figure groups the validated clips into score bands and compares the
share the experts found right (black, with a 95% interval) with what the
fitted curve predicts (red). Agreement means the curve describes the
data in that band.

![Calibration by score band: share found right against the fitted curve. Description: Two panels, nightjar and firefinch. Black points with intervals show the share of right clips per score band; a red line shows the fitted curve. For the nightjar the points sit close to the red line; the top band holds many clips and sits at 1.](figures/birds_calibration.png)

For the nightjar the points sit close to the curve in most bands, and
the top band (scores from `r f3(cal_top$conf_min)`, `r cal_top$n` clips)
has `r cal_top$n_correct` right clips against `r f3(cal_top$predicted)`
predicted. For the firefinch the shares are small and the intervals
wide, so there is no clear misfit but little else to learn.

### 6.2 AIC

I computed AIC (a score that compares models by fit, penalising extra
terms) for three models, as in the Wood & Kahl tutorial: no score, score
on the confidence scale, and score on the logit scale. I did not use it
to choose a model. We use only the logit-scale model. Difference from
the best model, per species:

```{r chunk-aic}
aic |>
  select(commonName, model, delta_aic) |>
  tidyr::pivot_wider(names_from = model, values_from = delta_aic) |>
  mutate(across(.cols = -commonName, .fns = f1)) |>
  select(
    Species = commonName,
    `No score` = null,
    `Confidence scale` = confidence_scale,
    `Logit scale` = logit_scale
  ) |>
  knitr::kable(format = "pipe", align = c("l", "r", "r", "r"))
```

For the nightjar the score clearly matters and the two scales are close
to tied. For the firefinch the score barely helps. For the plover the
comparison is not meaningful with one wrong clip.

### 6.3 What the checks show

They show that the nightjar curve fits the clips we have. They do not
show that it holds for other recorders, times or places; that is
section 7.

## 7. Sensitivity analysis: recorder, day-part, date

### 7.1 Why

Wood & Kahl warn that a threshold only holds for the conditions it was
validated under, but their tutorial does not run the check. I made it a
routine step. Device, date and hour come from the recording names.
Recorder model and sample rate are not in the data.

### 7.2 Recorders

For the nightjar, `r nj$n_neg` wrong clips sit on
`r dev_summary(species = sp_nightjar, what = "n_devices_with_negatives")`
of `r dev_summary(species = sp_nightjar, what = "n_devices")` recorders,
and `r pct0(dev_summary(species = sp_nightjar, what = "share_negatives_top2_devices"))`
of them are on two recorders (`r nj_top2[[1]]` and `r nj_top2[[2]]`),
which have no right clips. Recorder and score therefore cannot be
separated. The firefinch is the comparison: its wrong clips are spread
over `r dev_summary(species = sp_firefinch, what = "n_devices_with_negatives")`
recorders and the top two hold
`r pct0(dev_summary(species = sp_firefinch, what = "share_negatives_top2_devices"))`.

![Validated clips by recorder. Red marks the two recorders holding most of the species’ wrong clips. Description: Two panels, nightjar and firefinch, showing validated clips against confidence. Red points mark clips from the two recorders with most wrong clips; for the nightjar they cluster among the wrong clips at low scores.](figures/birds_device_effects.png)

### 7.3 Leave-one-recorder-out (a jackknife at recorder level)

I refit the nightjar threshold dropping each recorder in turn. It stays
between `r f2(min(nj_lodo$threshold))` and `r f2(max(nj_lodo$threshold))`
(full data `r f3(nj$glm_threshold)`). Dropping both of the two recorders
with most wrong clips gives `r f3(nj_lodo_pair$threshold)`, with
`r nj_lodo_pair$n_neg` wrong clips left.

### 7.4 Device-level bootstrap against clip bootstrap

Resampling recorders (with replacement) instead of clips gives
`r f2(nj_dboot$dboot_lower)` to `r f2(nj_dboot$dboot_upper)` for the
nightjar, against `r f2(nj_dboot$clip_boot_lower)` to
`r f2(nj_dboot$clip_boot_upper)` from resampling clips. The two agree,
so treating clips as independent does not make the interval look better
than it is. For the firefinch the recorder-level interval is
`r f2(ff_dboot$dboot_lower)` to `r f2(ff_dboot$dboot_upper)`, with
`r pct0(ff_dboot$dboot_share_failed)` of fits failing, and every
leave-one-out value stays between `r f3(min(ff_lodo$threshold))` and
`r f3(max(ff_lodo$threshold))`, outside the validated range.

### 7.5 Day-part

```{r chunk-daypart}
daypart |>
  mutate(delta_aic = if_else(is.na(delta_aic), "n/a", f3(delta_aic))) |>
  select(
    Species = commonName,
    `Day-part` = daypart,
    `Wrong clips` = n_neg,
    `Right clips` = n_pos,
    Estimable = estimable,
    `Delta AIC` = delta_aic
  ) |>
  knitr::kable(format = "pipe", align = c("l", "l", "r", "r", "l", "r"))
```

All `r dp_nj |> filter(n_neg > 0) |> pull(n_neg)` nightjar wrong clips are
from hours 16 to 23, and all `r dp_nj_night$n_pos` clips from hours 0 to
6 are right. A day-part effect cannot be estimated for the nightjar
because the night group has no wrong clips. Device, time of day and score
cannot be separated in this sample. Night looks cleaner, so applying the
evening-based threshold at night errs on the safe side, but that is
unverified. For the firefinch, adding a day-part term does not improve
the fit (delta AIC `r f1(dp_ff_dawn$delta_aic)`), though only
`r dp_ff_dawn$n_neg + dp_ff_dawn$n_pos` of its clips are dawn or night.

### 7.6 Date and season

Nightjar validated clips by week of the recording period:

```{r chunk-week}
coverage |>
  filter(commonName == sp_nightjar, dimension == "week") |>
  select(
    Week = level,
    Clips = n_clips,
    `Wrong clips` = n_neg,
    `Right clips` = n_pos
  ) |>
  knitr::kable(format = "pipe", align = c("l", "r", "r", "r"))
```

Date cannot be separated either. Nightjar wrong clips rise from
`r nj_week$wrong[[1]]` of `r nj_week$clips[[1]]` clips in week 1 to
`r nj_week$wrong[[3]]` of `r nj_week$clips[[3]]` in week 3, but
`r sum(nj_late$wrong_top2)` of the `r sum(nj_late$wrong)` wrong clips in
weeks 2 to 4 come from `r nj_top2[[1]]` and `r nj_top2[[2]]`, which were
validated only between `r fmt_day(nj_top2_dates[[1]])` and
`r fmt_day(nj_top2_dates[[2]])`. Among the other recorders, wrong clips
are rare in every week (`r knitr::combine_words(nj_other_by_week, and = " and ")`).
Date is a potential effect worth testing once validation covers more
weeks and recorders, but this sample cannot tell it apart from the
recorder effect.

Season is a separate reason for caution. Vocal activity is seasonal, and
breeding calls in particular do not happen all year, so the kind of sound
BirdNET scores can change through the year. All validated clips come from
`r n_days` days in `r format(period[[1]], "%B")` and
`r format(period[[2]], "%B %Y")`, a single season. A threshold fitted on
them may not hold at other times of year, which is why Wood & Kahl
advise another round of validation when the season changes.

### 7.7 Reading

The nightjar threshold is stable to dropping any one recorder and to
resampling recorders, so the number is stable. Its scope is not
established: it is estimated mostly from evening clips and from a few
recorders.

### 7.8 This should run on every refit

The recorder check, the device bootstrap and the day-part check are
cheap once the data exist, and a refit is exactly when new recorders,
hours and places enter the sample.

## 8. Planning for refits

When new validations arrive and thresholds are refitted, the same
diagnostics table (clip counts, range check, bootstrap failure share and
interval, calibration, recorder and hour checks) can feed a monitoring
dashboard for the senior data scientist and Biometrics. They can then
see at a glance whether a refit is trustworthy. The recorder and
environment sensitivity analysis is their job, run at project level on
every refit. There is one recorder per SD card, so the comparison is
across devices within a project.

## 9. Requests to the ornithologist and Biometrics

1.  **Random validation clips for every species**, drawn across more
    areas, recorders, hours of the day and seasons, and across the whole
    recording period for each recorder, so the sample captures the
    variation in each species’ calls and does not tie recorder, hour and
    date together. Species with no threshold (oriole, plover, firefinch)
    need this first.
2.  **Random nightjar clips at night.** Today every wrong nightjar clip
    is from the evening.
3.  **Metadata.** Tech should pass the recorder, BirdNET version and
    settings, site, habitat and date-time with every prediction. The
    full list is in the handoff to Tech (#5).

## 10. Limits

- **Scope of the nightjar threshold.** `r nj$n_neg` wrong clips,
  `r pct0(dev_summary(species = sp_nightjar, what = "share_negatives_top2_devices"))`
  on two recorders, all in the evening. It is labelled for all hours, with
  that evidence scope stated.
- **One grid and a short period.** All validated clips come from one
  grid over `r n_days` days (`r format(period[[1]])` to
  `r format(period[[2]])`). The grid is Grid2: every one of the
  predictions comes from it. The days sit in one season, so nothing in
  this sample tests whether the thresholds hold across the year.
- **Recall** is not equalised across species by a common precision
  target.

## 11. Observation labelling (#4)

### 11.1 The rule

Every prediction gets two new fields. `observation_status` is `observed`
when the species has a threshold and the confidence is at or above it,
`below_threshold` when the species has a threshold and the confidence is
below it, and `no_threshold` when the species has none. `observation`
holds the species name for `observed` rows and is empty otherwise. The
experts' validations do not override the model's label.

### 11.2 Result

```{r chunk-labelled}
labelled |>
  count(common_name, observation_status) |>
  tidyr::pivot_wider(
    names_from = observation_status,
    values_from = n,
    values_fill = 0L
  ) |>
  arrange(common_name) |>
  mutate(across(.cols = -common_name, .fns = big)) |>
  select(
    Species = common_name,
    observed,
    below_threshold,
    no_threshold
  ) |>
  knitr::kable(format = "pipe", align = c("l", "r", "r", "r"))
```

```{r labelled-totals}
n_status <- function(what) sum(labelled$observation_status == what)
share_status <- function(what) pct1(n_status(what = what) / n_predictions)
```

Of the `r big(n_predictions)` predictions, `r big(n_status(what = "observed"))`
(`r share_status(what = "observed")`) are observations,
`r big(n_status(what = "below_threshold"))`
(`r share_status(what = "below_threshold")`) are below threshold and
`r big(n_status(what = "no_threshold"))`
(`r share_status(what = "no_threshold")`) have no threshold. The labelled
file is `outputs/birds/birdnet_predictions_labelled.csv`: the 13 raw
columns plus the two new ones, one row per prediction.

### 11.3 Limits

Only the nightjar has observations. The oriole alone is
`r pct0(sum(labelled$common_name == sp_oriole) / n_predictions)` of all
predictions and has none until it is validated more widely (section 9).
The nightjar labels apply to all hours of the day, but its threshold was
checked only on evening clips from a few recorders (sections 7 and 10).

## Reproduce

The numbers in this report come from the files in `data/processed/`,
`outputs/birds/` and the raw predictions. The report is produced by
`rmarkdown::render("birds_report.Rmd")`. To regenerate and check the
numbers, run from the project root in a fresh R session:

```{r reproduce, echo=TRUE, eval=FALSE}
source(here::here("R", "birds", "01_load_validate.R"))
source(here::here("R", "birds", "02_threshold_evidence.R"))
source(here::here("R", "birds", "03_robustness_checks.R"))
source(here::here("R", "birds", "04_label_observations.R"))
testthat::test_dir(here::here("tests", "testthat"))
rmarkdown::render(here::here("birds_report.Rmd"))
```
