---
title: "BirdNET observations: handoff to the Tech team"
author: "Alban Sagouis"
date: "2026-10-08"
output:
  github_document:
    html_preview: false
    md_extensions: -implicit_figures
    pandoc_args: ["--markdown-headings=atx"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)

source(file = here::here("R", "birds", "config.R"))
paths <- birds_config$paths

# Every number in this handoff is read from data/processed/, outputs/birds/ or
# the config, written by R/birds/01 to 04. Nothing is typed by hand.
evidence <- readr::read_csv(file = paths$evidence, show_col_types = FALSE)
labelled <- readr::read_csv(
  file = paths$labelled,
  col_types = readr::cols(
    begin_time_s = readr::col_double(),
    confidence = readr::col_double(),
    .default = readr::col_character()
  )
) |>
  mutate(
    # Stable key of the current file; the ML pipeline owns the real prediction_id
    prediction_key = paste(begin_path, begin_time_s, species_code, sep = "|")
  )

f2 <- function(x) sprintf(fmt = "%.2f", x)
f3 <- function(x) sprintf(fmt = "%.3f", x)
f4 <- function(x) sprintf(fmt = "%.4f", x)
f17 <- function(x) format(x, digits = 17, scientific = FALSE)
pct <- function(x) sprintf(fmt = "%.0f%%", 100 * x)
big <- function(x) format(x, big.mark = ",", trim = TRUE)

n_pred <- nrow(labelled)
n_species <- dplyr::n_distinct(labelled$species_code)
n_key_dup <- sum(duplicated(labelled$prediction_key))
n_observed <- sum(labelled$observation_status == "observed")

# One row per species: code, status, cutoff and the labels it got
species_map <- labelled |>
  distinct(common_name, species_code)
thr <- evidence |>
  inner_join(
    y = species_map,
    by = join_by(commonName == common_name),
    relationship = "one-to-one",
    unmatched = c(x = "error", y = "error")
  )
status_counts <- labelled |>
  count(species_code, observation_status) |>
  tidyr::pivot_wider(
    names_from = observation_status,
    values_from = n,
    values_fill = 0L
  )
species_tbl <- thr |>
  inner_join(
    y = status_counts,
    by = join_by(species_code),
    relationship = "one-to-one",
    unmatched = c(x = "error", y = "error")
  ) |>
  mutate(
    n_total = observed + below_threshold + no_threshold,
    share_observed = observed / n_total
  )

nj <- thr |> filter(status == "fitted threshold")
stopifnot(nrow(nj) == 1L)
nj_cutoff <- nj$labelling_cutoff
nj_code <- nj$species_code
no_thr <- thr |> filter(status == "no_threshold")
n_fitted <- sum(thr$status == "fitted threshold")
n_none <- nrow(no_thr)
cfg <- birds_config
no_match_reason <- "no thresholds row for this species, BirdNET version and settings"
versions <- unique(readr::read_csv(
  file = paths$validations,
  col_select = "vBirdNET",
  show_col_types = FALSE
)$vBirdNET)
```

## TL;DR

- **What to build.** A labelling step that runs after the ML predictions are stored: it joins
  every prediction to a thresholds table and writes one row per prediction to an observations
  table. It is one SQL statement (`sql/birds_label_observations.sql`) and four tables
  (`sql/birds_schema.sql`). No statistics run at runtime.
- **The key rule.** `confidence >= cutoff` gives `observed`, a smaller confidence gives
  `below_threshold`, and a missing cutoff or missing thresholds row gives `no_threshold`.
  `observation` holds the species name only when the status is `observed`. In the current data
  `r big(n_observed)` of `r big(n_pred)` predictions are observed.
- **What Tech must do.** Create the tables, run the statement after every upload, store the
  metadata listed below with every prediction, make the validation export carry the prediction
  id (data-issue #5), and expose the monitoring numbers.
- **What is not Tech's decision.** Thresholds. Biometrics inserts rows into the thresholds
  table; Tech reads them and never edits one. `r n_none` of `r nrow(thr)` species currently
  have no threshold, which is a normal state the pipeline handles, not an error.

## Inputs and outputs

### Inputs

- `birdnet_predictions`: one row per BirdNET detection, written by the existing ML pipeline.
  The current file has `r big(n_pred)` rows for `r n_species` species and the columns
  `r paste(names(readr::read_csv(file = paths$predictions, n_max = 0, show_col_types = FALSE)), collapse = ", ")`.
- `birdnet_thresholds`: one row per species, BirdNET version and settings, written by
  Biometrics. Today `r n_fitted` row holds a fitted threshold and `r n_none` rows hold none.
- `birdnet_validations`: expert verdicts on predictions. Used offline by Biometrics for refits;
  Tech only stores them.

### Outputs

- `birdnet_observations`: one row per prediction with `observation_status`, `observation`,
  the `threshold_id` used and a reason when there is no label.

The table definitions are in the next section and in `sql/birds_schema.sql`.

## The labelling contract

### How observations are derived from predictions

A prediction is a detection with a BirdNET confidence. It becomes an observation only if its
confidence reaches the cutoff for that species, the BirdNET version and the settings that
produced it. The cutoff is the confidence at which the validated clips of that species reach
`r pct(cfg$target_precision)` precision. Biometrics estimates it offline; at runtime it is just a number in a table.

The rule is applied in this order, and the first line that matches decides:

1. No thresholds row matches the prediction's species, BirdNET version and settings, or the
   matching row has `cutoff` NULL: `observation_status = 'no_threshold'`.
2. `confidence >= cutoff`: `observation_status = 'observed'`.
3. Otherwise: `observation_status = 'below_threshold'`.

### The three statuses

| `observation_status` | Fixed meaning | `observation` |
|:---|:---|:---|
| `observed` | The species is recorded as present at this time and place. | Species common name |
| `below_threshold` | A cutoff exists and the confidence is under it. Not an observation. | NULL |
| `no_threshold` | No cutoff exists for this species, version and settings. Not an observation, not a rejection. | NULL |

Thresholds rows have a `status` of either `fitted threshold` (cutoff present) or
`no_threshold` (cutoff NULL, with a `reason`). Tech does not interpret the reason; it copies it
to `status_reason` in the observations table.

### Worked example

Real rows from the current data, labelled with the real cutoff for
`r nj$commonName` (`r nj_code`): `r f4(nj_cutoff)`.

```{r worked-example}
below <- labelled |>
  filter(species_code == nj_code, confidence < nj_cutoff) |>
  slice_max(order_by = confidence, n = 1, with_ties = FALSE)
above <- labelled |>
  filter(species_code == nj_code, confidence >= nj_cutoff) |>
  slice_min(order_by = confidence, n = 1, with_ties = FALSE)
top <- labelled |>
  filter(species_code == nj_code) |>
  slice_max(order_by = confidence, n = 1, with_ties = FALSE)
stopifnot(top$confidence == 1)
others <- no_thr |>
  pull(species_code) |>
  lapply(FUN = function(code) {
    labelled |>
      filter(species_code == code) |>
      slice_max(order_by = confidence, n = 1, with_ties = FALSE)
  }) |>
  bind_rows()
example <- bind_rows(below, above, top, others) |>
  transmute(
    prediction_key,
    species_code,
    confidence,
    cutoff = ifelse(
      test = is.element(species_code, nj_code),
      yes = f4(nj_cutoff),
      no = "none"
    ),
    observation_status,
    observation = ifelse(is.na(observation), "NULL", observation)
  )
knitr::kable(x = example, align = "llrrll")
```

Rows 1 and 2 are the nightjar predictions closest to the cutoff on either side. Row 3 is the
highest nightjar confidence, which is exactly 1: no logit is computed at runtime, so 1.0 needs
no special handling. The remaining rows are each other species' highest-confidence prediction.
They still get `no_threshold`, however confident BirdNET is, because no threshold has been
established for them.

## Tables

### Schema

The definitions are in `sql/birds_schema.sql` (PostgreSQL 14 or later) and reproduced here.

```{r schema-sql, results="asis"}
cat("```sql\n")
cat(readLines(con = here::here("sql", "birds_schema.sql")), sep = "\n")
cat("\n```\n")
```

### Notes on the tables

- `prediction_id` must be assigned by the ML pipeline and must be the same on every re-ingest
  of the same detection. In the current file, recording path + start time + species code is a
  unique key (`r n_key_dup` duplicates in `r big(n_pred)` rows) and can seed it.
- `birdnet_thresholds` is insert-only for Biometrics. A refit adds rows; it never changes old
  ones. The check constraint ties `cutoff` to `status` so a row cannot say "fitted" without a
  number.
- `settings` and `birdnet_settings` are compared for equality, so both must be written with the
  same keys by the same code.

## The runtime labelling step

`sql/birds_label_observations.sql` is one `INSERT ... SELECT ... ON CONFLICT` statement.

```{r label-sql, results="asis"}
cat("```sql\n")
cat(readLines(con = here::here("sql", "birds_label_observations.sql")), sep = "\n")
cat("\n```\n")
```

- It picks the current thresholds row per species, BirdNET version and settings: the latest
  `valid_from` that has already started.
- It is idempotent. `prediction_id` is the primary key of `birdnet_observations`, so a rerun
  with unchanged thresholds updates nothing (the `WHERE` on the conflict clause skips rows
  whose `threshold_id` is unchanged).
- After Biometrics inserts a new thresholds row, the next run relabels exactly the predictions
  whose row changed and stamps the new `threshold_id`. Run it after every upload and after
  every refit; no other trigger is needed.
- It contains no statistics and no logit. Run it in one transaction.

## Refitting

Thresholds are refitted offline by Biometrics, never by the platform. A refit is triggered by
new validations or by a new BirdNET version (current validations are from BirdNET
`r paste(versions, collapse = ", ")`), is versioned by BirdNET version and settings, and ends
with Biometrics inserting rows into `birdnet_thresholds`. Tech does nothing for a refit except
rerun the labelling statement; the table never has rows edited or deleted by Tech.

## Configuration Biometrics can change without code

These values live in `R/birds/config.R`, are used only by the offline refit and are never read
at runtime. Changing one produces new thresholds rows, not a code change on the platform.

```{r config-table}
config_tbl <- tibble::tibble(
  value = c(
    "target_precision",
    "min_smaller_group",
    "n_boot",
    "ci_level"
  ),
  current = c(
    cfg$target_precision,
    cfg$min_smaller_group,
    cfg$n_boot,
    cfg$ci_level
  ),
  meaning = c(
    "Precision the cutoff must reach on the validated clips",
    "Fewest clips the smaller outcome group (right or wrong) may have for a fit",
    "Bootstrap resamples for the interval and failure share",
    "Level of the reported interval (`ci_low`, `ci_high`)"
  )
)
knitr::kable(x = config_tbl)
```

The range gate (the fitted threshold must lie inside the validated score range) is applied to
every species on every refit and has no switch; its result is the `in_range` column.

## Edge cases and failure behaviour

| Case | Behaviour |
|:---|:---|
| Species with no thresholds row | `no_threshold`, `threshold_id` NULL, reason says no row matches. Not an error. |
| Matching row with `cutoff` NULL | `no_threshold`, `threshold_id` of that row, its `reason` copied. |
| Confidence exactly 1.0 | Compared like any other value. No logit at runtime, so nothing can overflow. |
| Confidence equals the cutoff | `observed` (the rule is `>=`). |
| New BirdNET version or different settings | No row matches, so `no_threshold` with a reason. Labels return when Biometrics inserts rows for the new version. |
| The statement is run twice | No change. |
| Same detection ingested twice | The primary key on `prediction_id` rejects the duplicate; the ML pipeline should use `ON CONFLICT DO NOTHING`. |
| Validation arrives late | It is stored and used at the next refit. Labels change only when a new thresholds row is inserted. |
| Row with confidence outside 0 to 1 | Rejected by the check constraint on insert. Fix at the source. |

## Metadata to store with every prediction

Store device, recorder model, sample rate, BirdNET version and settings (including sensitivity
and overlap), site, habitat and datetime, so that Biometrics can check at refit time whether a
threshold holds across recorders, sites and times, which cannot be recovered afterwards. Tech
only stores and exposes these fields; the analysis is not Tech's. The columns are in
`birdnet_predictions`.

## Monitoring

### Per batch

The share of predictions labelled `observed`, per species and batch. A sudden change usually
means a new BirdNET version or settings (all `no_threshold`) or a broken upload.

```sql
SELECT p.project_id,
       p.ingested_at::date AS batch_date,
       p.species_code,
       count(*) AS n_predictions,
       avg((o.observation_status = 'observed')::int) AS share_observed,
       avg((o.observation_status = 'no_threshold')::int) AS share_no_threshold
FROM birdnet_observations AS o
JOIN birdnet_predictions AS p USING (prediction_id)
GROUP BY p.project_id, p.ingested_at::date, p.species_code;
```

Current values in the example data:

```{r monitoring-table}
species_tbl |>
  transmute(
    species_code,
    status,
    n_predictions = n_total,
    observed,
    share_observed = pct(share_observed)
  ) |>
  knitr::kable(align = "llrrr")
```

### Refit monitoring

A Biometrics dashboard reads the diagnostic columns of `birdnet_thresholds`: `n_negative`,
`n_positive`, `smaller_group`, `in_range`, `boot_share_failed`, `ci_low`, `ci_high`, `status`
and `reason`. Tech exposes the table to the dashboard; nothing else is needed.

## Upstream data issue

Validations cannot be linked reliably to predictions today. The validation export must carry
the prediction id (or the clip start position), a recording id shared with the predictions, and
full-precision scores. Details and the evidence are in data-issue #5. This blocks
`birdnet_validations.prediction_id`, so it matters for refits, not for runtime labelling.

## Acceptance tests

Each row is an input prediction and its expected output, derived from the real data and the
real `r nj$commonName` cutoff (`r f17(nj_cutoff)`). Run with one thresholds row for the
nightjar (`fitted threshold`, `birdnet_version` fixed, settings fixed) plus the rows for the
other species in the table below.

```{r acceptance-tests}
other_thr <- no_thr |>
  select(species_code, reason)
tests <- bind_rows(
  below |> mutate(case = "Real nightjar, just below the cutoff"),
  above |> mutate(case = "Real nightjar, just above the cutoff"),
  above |>
    mutate(confidence = nj_cutoff, case = "Nightjar, confidence equal to the cutoff (constructed)"),
  top |> mutate(case = "Real nightjar, confidence 1.0"),
  others |>
    mutate(case = paste("Real", common_name, "prediction, no threshold")),
  top |>
    mutate(
      birdnet_version = "vNEW",
      case = "Nightjar from an unseen BirdNET version (constructed)"
    )
) |>
  mutate(
    birdnet_version = coalesce(birdnet_version, versions[[1]]),
    expected_status = case_when(
      birdnet_version != versions[[1]] ~ "no_threshold",
      is.element(species_code, nj_code) & confidence >= nj_cutoff ~ "observed",
      is.element(species_code, nj_code) ~ "below_threshold",
      .default = "no_threshold"
    ),
    expected_threshold = case_when(
      birdnet_version != versions[[1]] ~ "NULL",
      is.element(species_code, nj_code) ~ "nightjar row",
      .default = "that species' row"
    )
  ) |>
  left_join(
    y = other_thr,
    by = join_by(species_code),
    relationship = "many-to-one",
    unmatched = "drop"
  ) |>
  mutate(
    expected_reason = case_when(
      birdnet_version != versions[[1]] ~ no_match_reason,
      is.element(species_code, nj_code) ~ "NULL",
      .default = reason
    ),
    expected_observation = ifelse(
      test = expected_status == "observed",
      yes = common_name,
      no = "NULL"
    )
  )
tests |>
  transmute(
    n = row_number(),
    case,
    species_code,
    confidence = ifelse(
      test = confidence == nj_cutoff,
      yes = f17(confidence),
      no = as.character(confidence)
    ),
    expected_status,
    expected_observation,
    expected_threshold,
    expected_reason
  ) |>
  knitr::kable(align = "rllrllll")
```

Also test: running the statement twice leaves `labelled_at` unchanged on every row; inserting
a later `valid_from` row for the nightjar relabels only nightjar predictions and changes their
`threshold_id`; a row with confidence 1.5 is rejected on insert.

## Versioning

- This document and both SQL files are versioned in git; a schema change is a migration that
  only adds columns.
- Thresholds are versioned by their rows: `birdnet_version`, `settings`, `fitted_at` and
  `valid_from`. Old rows stay.
- Every observation stores the `threshold_id` it was labelled with, so any label can be traced
  to the exact row and the statement that produced it.

## Stack

Python 3.13, PostgreSQL, Redis and Docker Compose on Azure, as already in use; this handoff
adds tables and one SQL statement and prescribes nothing about the deployment.
