---
title: "Herbaceous vegetation survey: errors and warnings for the field teams"
author: "Alban Sagouis"
date: "2026-10-08"
output:
  github_document:
    html_preview: false
    md_extensions: -implicit_figures
    pandoc_args: ["--markdown-headings=atx"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)

source(file = here::here("R", "vegetation", "config.R"))

# Every number comes from the two CSV files written by R/vegetation/02_run_checks.R
issues <- readr::read_csv(
  file = veg_config$paths$field_issues,
  col_types = readr::cols(.default = readr::col_character())
)
flags <- readr::read_csv(
  file = veg_config$paths$flags,
  col_types = readr::cols(.default = readr::col_character())
)
catalogue <- readr::read_csv(
  file = veg_config$paths$catalogue,
  col_types = readr::cols(.default = readr::col_character())
)

n_plots <- n_distinct(issues$plot)
n_submissions <- n_distinct(issues$survey_key)
# Submissions come from the staged survey table (written by R/vegetation/01_load_join.R)
survey <- readr::read_csv(
  file = veg_config$paths$survey,
  col_types = readr::cols(.default = readr::col_character())
)
rejected_keys <- veg_rejected_keys(survey = survey)
n_rejected <- length(rejected_keys)
n_surveys <- nrow(survey) - n_rejected
n_rejected_issues <- sum(
  is.element(flags$severity, c("error", "warning")) & is.element(flags$survey_key, rejected_keys)
)
n_error <- sum(issues$severity == "error")
n_warning <- sum(issues$severity == "warning")
n_field <- sum(issues$who_can_resolve == "field team")
n_manager <- sum(issues$who_can_resolve == "data manager")
n_tech <- sum(issues$who_can_resolve == "Tech")
n_errors_subs <- n_distinct(issues$survey_key[issues$severity == "error"])
n_info <- sum(flags$severity == "info")
```

## TL;DR

- This list has `r nrow(issues)` items in `r n_submissions` of the `r n_surveys` accepted submissions
  (`r n_plots` plots): `r n_error` errors and `r n_warning` warnings. Nothing was changed or removed in the data.
- `r n_rejected` submissions were rejected in ODK and are not on this list; their `r n_rejected_issues`
  errors and warnings stay in `outputs/vegetation/flags.csv`.
- Errors break a rule of the survey protocol (SOP) or contradict other answers. Warnings need a look and
  may be fine.
- `r n_field` items can be resolved by the field team, `r n_manager` by the data manager and `r n_tech`
  by the Tech team (they come from the form, not from the field).
- The most common field problems: species names typed with a space instead of an underscore
  (`r sum(issues$check_id == "SPE-02")` names), quadrat locations far from the plot midpoint
  (`r sum(issues$check_id == "SPA-04")` quadrats), and extra species announced but not recorded
  (`r sum(issues$check_id == "CON-04")` quadrats).
- `r n_info` further notes (for example species still to be identified) are in `outputs/vegetation/flags.csv`
  and are not repeated here.

## How to read this list

Each plot has one section and each submission (one visit) one sub-section. Quadrat numbers are the numbers
in the form. "Check" lines say what to look at. Section numbers refer to the Herbaceous Vegetation Surveys
SOP and the Vegetation Plot Registration SOP (version 2026.1). The full list as a table is
`outputs/vegetation/issues_for_field_teams.csv`; the rules behind each item are in
`outputs/vegetation/check_catalogue.csv`.

```{r sections, results='asis'}
# One bullet per issue, then what to check and the SOP sections once per submission
cat_submission <- function(rows) {
  rows <- arrange(.data = rows, match(x = severity, table = c("error", "warning")), check_id, as.integer(quadrat))
  cat(sprintf(
    "### %s, recorded by %s (%d item%s)\n\n",
    rows$date[1],
    ifelse(test = is.na(rows$recorder[1]), yes = "unknown recorder", no = rows$recorder[1]),
    nrow(rows),
    ifelse(test = nrow(rows) == 1, yes = "", no = "s")
  ))
  label <- ifelse(test = rows$severity == "error", yes = "Error", no = "Warning")
  cat(paste0("- ", label, ": ", rows$message, " (", rows$who_can_resolve, ")"), sep = "\n")
  cat("\n")
  cat("Check:\n\n")
  cat(paste0("- ", unique(rows$what_to_check)), sep = "\n")
  cat("\n")
  cat("SOP: ", paste(unique(rows$sop_section), collapse = "; "), "\n\n", sep = "")
}

for (plot in sort(unique(issues$plot))) {
  plot_rows <- filter(.data = issues, plot == .env$plot)
  cat(sprintf("## %s\n\n", plot))
  for (key in unique(plot_rows$survey_key)) {
    cat_submission(rows = filter(.data = plot_rows, survey_key == .env$key))
  }
}
```
