---
title: "Herbaceous vegetation: survey QA/QC"
format:
  dashboard:
    embed-resources: true
    theme: custom.scss
    nav-buttons:
      - icon: github
        href: https://github.com/albansagouis/naturalstate-take-home
brand: _brand.yml
---

```{r}
#| label: setup
#| include: false
# The dashboard only reads the files the report is built from (outputs/vegetation/,
# figures/). It runs no analysis.
library(dplyr)
options(knitr.kable.NA = "")
source(file = here::here("R", "vegetation", "config.R"))
paths <- veg_config$paths
read_out <- function(path) readr::read_csv(file = path, show_col_types = FALSE)

survey <- read_out(path = paths$survey_summary)
plots <- read_out(path = paths$plot_summary)
totals <- read_out(path = paths$totals) |> filter(version == "accepted")
locations <- read_out(path = paths$plot_locations)
flags <- read_out(path = paths$flags)
effort <- read_out(path = paths$effort)
curves <- read_out(path = paths$accumulation_curves)

html_table <- function(data) {
  htmltools::HTML(knitr::kable(x = data, format = "html", table.attr = 'class="table table-sm table-striped"'))
}
f1 <- function(x) sprintf(fmt = "%.1f", x)
f2 <- function(x) sprintf(fmt = "%.2f", x)
```

# Summaries

## Row {height=24%}

```{r}
#| label: box-surveys
#| content: valuebox
#| title: "Accepted surveys"
list(icon = "clipboard-check", color = "primary", value = totals$n_surveys)
```

```{r}
#| label: box-rejected
#| content: valuebox
#| title: "Rejected in ODK (not counted)"
list(icon = "x-circle", color = "secondary", value = sum(survey$review_state == "rejected", na.rm = TRUE))
```

```{r}
#| label: box-plots
#| content: valuebox
#| title: "Plots surveyed"
list(icon = "geo-alt", color = "primary", value = paste(totals$n_plots_surveyed, "of", totals$n_plots_registered))
```

```{r}
#| label: box-quadrats
#| content: valuebox
#| title: "Quadrats"
list(icon = "grid-3x3", color = "primary", value = totals$n_quadrats)
```

```{r}
#| label: box-taxa
#| content: valuebox
#| title: "Identified taxa"
list(icon = "flower1", color = "success", value = totals$richness_all_plots)
```

```{r}
#| label: box-flags
#| content: valuebox
#| title: "Flags E / W / I"
list(icon = "flag", color = "warning", value = paste(totals$n_flags_error, totals$n_flags_warning, totals$n_flags_info, sep = "/"))
```

## Row {height=64%}

### Column {.tabset}

```{r}
#| label: survey-summary
#| title: "Survey summary"
survey |>
  mutate(plot = sub(pattern = "^SavMon_LW_", replacement = "", x = plot_name), duration_min = f1(duration_min)) |>
  select(plot, status = plot_status, review = review_state, date = survey_date, recorder,
    `duration (min)` = duration_min, n_quadrats, taxa = richness, unknowns = n_unknown_labels,
    errors = n_error, warnings = n_warning, info = n_info) |>
  html_table()
```

```{r}
#| label: plot-summary
#| title: "Plot summary"
plots |>
  mutate(plot = sub(pattern = "^SavMon_LW_", replacement = "", x = plot_name)) |>
  select(plot, status = plot_status, surveyed, n_quadrats, taxa = gamma_richness,
    unknowns = n_unknown_labels, errors = n_error, warnings = n_warning) |>
  html_table()
```

## Row {height=12%}

All numbers are the report's (`outputs/vegetation/`); the dashboard adds no analysis. The Natural State
logo and colours are used for this mock-up only.

# Map

```{r}
#| label: map
#| title: "Plot locations coloured by worst flag severity"
library(leaflet)
drawn <- locations |>
  filter(!is.na(lon), !is.na(lat)) |>
  mutate(
    label = sub(pattern = "^SavMon_LW_", replacement = "", x = plot_name),
    colour = unname(veg_config$severity_colours[worst_severity]),
    outline = if_else(condition = plot_status == "backup", true = veg_config$brand[["navy"]], false = "#0B0809"),
    width = if_else(condition = plot_status == "backup", true = 5, false = 2),
    popup = sprintf(
      fmt = "<b>%s</b>%s<br>Worst flag: %s<br>%s errors, %s warnings, %s info<br>%s",
      plot_name,
      if_else(condition = plot_status == "backup", true = " (backup plot)", false = ""),
      worst_severity, n_error, n_warning, n_info,
      if_else(condition = surveyed, true = "Surveyed", false = "Registered, not surveyed")
    )
  )
# fadeAnimation off: tiles appear at full strength at once (also in the screenshot)
leaflet(data = drawn, options = leafletOptions(fadeAnimation = FALSE, zoomAnimation = FALSE)) |>
  # Satellite imagery (Esri World Imagery) needs an internet connection and a page served
  # over http(s); opened offline or from disk the markers still draw, without tiles.
  addProviderTiles(provider = providers$Esri.WorldImagery) |>
  addCircleMarkers(
    lng = ~lon, lat = ~lat, radius = 9, weight = ~width, color = ~outline,
    fillColor = ~colour, fillOpacity = 0.95, popup = ~popup, label = ~label
  ) |>
  addLegend(
    position = "bottomright", title = "Worst flag<br>(navy rim = backup plot)",
    colors = unname(veg_config$severity_colours[veg_config$severity_levels[1:3]]),
    labels = c("Error", "Warning", "Info only"), opacity = 1
  )
```

# Flags

```{r}
#| label: flags-table
#| results: asis
#| title: "Flags (filter by severity, check or plot)"
# A small self-contained filter: DT and reactable are not installed.
esc <- htmltools::htmlEscape
options_of <- function(x) paste0('<option value="', esc(x), '">', esc(x), "</option>", collapse = "")
rows <- flags |>
  mutate(plot = sub(pattern = "^SavMon_LW_", replacement = "", x = plot_name)) |>
  arrange(match(x = severity, table = veg_config$severity_levels), check_id, plot)
body <- sprintf(
  fmt = '<tr data-sev="%s" data-chk="%s" data-plot="%s"><td><span style="background:%s;color:%s;padding:1px 6px;border-radius:3px">%s</span></td><td>%s</td><td>%s</td><td>%s</td><td>%s</td><td>%s</td></tr>',
  rows$severity, rows$check_id, esc(rows$plot),
  unname(veg_config$severity_colours[rows$severity]),
  if_else(condition = rows$severity == "warning", true = "#0B0809", false = "#FFFFFF"),
  rows$severity, rows$check_id, esc(rows$plot), rows$survey_date, esc(rows$message), esc(rows$who_can_resolve)
)
# A raw HTML block: pandoc must not parse the 960 table rows as Markdown (it did not finish)
cat(
  "```{=html}\n",
  '<div style="display:flex;flex-direction:row !important;gap:16px;flex-wrap:wrap;align-items:center;justify-content:flex-start;flex:0 0 auto;padding:6px 0 10px">',
  '<label>Severity <select id="f-sev"><option value="">all</option>', options_of(unique(rows$severity)), "</select></label>",
  '<label>Check <select id="f-chk"><option value="">all</option>', options_of(sort(unique(rows$check_id))), "</select></label>",
  '<label>Plot <select id="f-plot"><option value="">all</option>', options_of(sort(unique(rows$plot))), "</select></label>",
  '<span id="f-count"></span></div>',
  '<table class="table table-sm table-striped" id="flags-tbl"><thead><tr><th>Severity</th><th>Check</th><th>Plot</th><th>Date</th><th>Message</th><th>Who resolves</th></tr></thead><tbody>',
  paste(body, collapse = ""), "</tbody></table>",
  "<script>(function(){var s=['f-sev','f-chk','f-plot'].map(function(i){return document.getElementById(i)});",
  "var rows=document.querySelectorAll('#flags-tbl tbody tr');",
  "function run(){var n=0;rows.forEach(function(r){var ok=(!s[0].value||r.dataset.sev===s[0].value)&&(!s[1].value||r.dataset.chk===s[1].value)&&(!s[2].value||r.dataset.plot===s[2].value);r.style.display=ok?'':'none';if(ok)n++;});",
  "document.getElementById('f-count').textContent=n+' of '+rows.length+' flags';}",
  "s.forEach(function(e){e.addEventListener('change',run)});run();})();</script>",
  "\n```\n",
  sep = ""
)
```

# Sampling effort

## Row {height=22%}

```{r}
#| label: box-effort-full
#| content: valuebox
#| title: "Plots with 20 quadrats"
list(icon = "grid-3x3", color = "primary", value = paste(sum(effort$n_quadrats >= veg_config$expected_quadrats_per_plot), "of", nrow(effort)))
```

```{r}
#| label: box-effort-asymptote
#| content: valuebox
#| title: "Plots approaching an asymptote"
list(icon = "graph-up", color = "success", value = paste(sum(effort$approaches_asymptote), "of", nrow(effort)))
```

```{r}
#| label: box-effort-completeness
#| content: valuebox
#| title: "Median completeness (observed / Chao2)"
list(icon = "percent", color = "warning", value = sprintf(fmt = "%.0f%%", 100 * median(effort$completeness)))
```

## Row {height=66%}

```{r}
#| label: effort-table
#| title: "Effort per plot"
effort |>
  arrange(completeness) |>
  mutate(
    plot = sub(pattern = "^SavMon_LW_", replacement = "", x = plot_name),
    `Chao2 (95% interval)` = sprintf(fmt = "%.1f (%.1f to %.1f)", estimated_richness, estimated_lcl, estimated_ucl),
    completeness = sprintf(fmt = "%.0f%%", 100 * completeness),
    coverage = f2(sample_coverage),
    asymptote = if_else(condition = approaches_asymptote, true = "yes", false = "no")
  ) |>
  select(plot, quadrats = n_quadrats, observed = observed_richness, `Chao2 (95% interval)`,
    completeness, coverage, asymptote) |>
  html_table()
```

```{r}
#| label: effort-curves
#| title: "Species accumulation curves"
library(ggplot2)
curves |>
  mutate(plot = sub(pattern = "^SavMon_LW_Plot_", replacement = "", x = plot_name)) |>
  ggplot(mapping = aes(x = n_quadrats, y = richness)) +
  geom_ribbon(mapping = aes(ymin = pmax(richness - sd, 0), ymax = richness + sd), fill = "#A5D3B8", alpha = 0.6) +
  geom_line(colour = "#4A5E9A", linewidth = 0.6) +
  facet_wrap(facets = vars(plot), ncol = 6) +
  labs(x = "Quadrats added (random order)", y = "Identified taxa (mean)") +
  veg_config$theme
```

## Row {height=12%}

Completeness is observed taxa divided by the Chao2 estimate; a plot approaches an asymptote at completeness
of at least `r veg_config$completeness_min` and sample coverage of at least `r veg_config$coverage_min`
(provisional thresholds, to confirm with Natural State). The ribbon is one standard deviation over the random orderings.
